# Properties of Estimators and Inference

## Properties of point estimators

This exercise is partly based on Exercise 6.1 of the book `Statistical Rethinking` by M.D. Edge. We'll draw many simulated samples from a known distribution with known parameters. We will then pretend this are instances of real datasets and estimate some parameters of the distributions using different estimators. Our goal will be to analyze properties of these parameter estimators, namely their bias and variance. Recall the definitions of these two properties:

Bias: $$B(\hat\theta_n) =  E[\hat\theta_n(D) - \theta] $$

The bias reflects the average difference between our estimator and the true parameter.

Variance: $$Var(\hat\theta_n) =  E[ (\hat\theta_n(D) - E[\hat\theta_n(D)] )^2] $$

The variance reflects the variation of the estimator around its own mean (regardless of what the true parameter is).

**Note**: in real life, we will generally not know the distribution from which our data comes from, or the value of the parameters of that distribution (otherwise we wouldn't be performing statistical inference). When developing or applying a particular estimator to our data, it is thus important to test the properties of that estimator under a variety of simulations and parameter choices. These simulations should aim to, as much as possible, mimic the possible range of processes and parameters that could be generating our data.


We will first create a function that draws a user-specified number (`nsim`) of independent data samples of size `sampsize` from a distribution of choice. By default this distribution is the Normal distribution (`rnorm`), but it can be changed by the user:

```{r}
simsamps <- function(sampsize, nsim = 10000, rsim = rnorm, ...){
  sims <- rsim(sampsize*nsim,...) # generate draws from a specified distribution
  simmat <- matrix(sample(sims), nrow=nsim, ncol=sampsize) # organize those draws into a matrix
  return(simmat) # provide the matrix as output
}
```

The output of this function is a matrix, with row number equal to the number of generated simulations, and column number equal to the size of each simulation. For example, to create 10,000 simulations, each of size 30, from a normal distribution, we would write:

```{r}
testmat <- simsamps(30,10000,rnorm) # generate matrix of 10,000 Normal data sets, each of size 30
dim(testmat) # dimensions of matrix
```

What type of Normal distribution are we sampling from though? By default, the function rnorm samples from a standard Normal distribution with mean equal to 0 and standard deviation equal to 1. In our function `simsamps`, we are only feeding one argument to the normal distribution (the number of draws we want to obtain from it). The three dots placed in both the argument of the `simsamps` function and in its internal call to `rsim` allows us to feed more parameters to `rsim`:

```{r}
testmat <- simsamps(30,100,rnorm,mean=3,sd=5) # generate matrix of 1000 Normal(3,5) data sets of size 30
```

We can obtain the sample mean and sample median from each of these datasets, using the `apply` function:

```{r}
testmean <- apply(testmat,1,mean)
testmedian <- apply(testmat,1,median)
```

**Exercise**: Create 20,000 simulated data sets of size 5, each drawn from a Normal distribution with expected value equal to 4 and standard deviation equal to 10. Create a histogram of the sample means from all the data sets. Draw a red line where the expected value of the original source distribution is located. Next, do the same for all the sample medians.

Now, try answering these questions: Is the sample mean a biased estimator of the expected value of a normal distribution? Is the sample median a biased estimator of the expected value of a normal distribution? Which estimator has the highest variance? Which estimator would you use if given a dataset like one of the ones we simulated, in order to estimate the expected value of this Normal distribution?

```{r, eval=FALSE, echo=FALSE}
EV <- 4
SD <- 10
normmat <- simsamps(5,20000,rnorm,mean=EV,sd=SD) 
normmean <- apply(normmat,1,mean)
normmedian <- apply(normmat,1,median)
par(mfrow=c(2,1))
hist(normmean,xlim=c(-25,25))
abline(v=mean,col="red")
hist(normmedian,xlim=c(-25,25))
abline(v=mean,col="red")
par(mfrow=c(1,1))
```

**Exercise**: Create 20,000 simulated data sets of size 10, each drawn from an Exponential distribution with rate equal to 2. Create a histogram of the sample means from all the data sets. Draw a red line where the expected value of the original source distribution is located.Recall that the expected value of this distribution is 1/rate = 0.5.

Now, try answering these questions: Is the sample mean a biased estimator of the expected value of an exponential distribution? Is the sample median a biased estimator of the expected value of an exponential distribution? Which of these two estimators would you use if you were trying to estimate the rate of an exponential process (e.g. the rate at which buses arrive at a station), from a dataset like one of the ones we just simulated?

```{r,eval=FALSE, echo=FALSE}
exprate <- 2
expmat <- simsamps(10,20000,rexp,rate=exprate) 
expmean <- apply(expmat,1,mean)
expmedian <- apply(expmat,1,median)
par(mfrow=c(1,2))
hist(expmean)
abline(v=1/2,col="red")
hist(expmedian)
abline(v=1/2,col="red")
par(mfrow=c(1,1))
```

## Building confidence intervals

This exercise is partly based on Exercise 7.2 of the book `Statistical Rethinking` by M.D. Edge.

## ROC curve