# Resampling

Resampling consists in **re-using** our data several times in smart ways, particularly when one doesn't want to make too many assumptions about the underlying distribution that is generating our data. There are many reasons for which one would want to re-use one's data, and today we'll look at three different cases in which resampling can help us:

1. Estimating confidence intervals (using bootstrapping)

2. Testing a model (using permutation)

3. Evaluating model fit (using cross-validation)

For these exercises, we'll use the `tidyverse` library, so let's load it first.

```{r, message=FALSE}
library("tidyverse")
```

## The bootstrap


We’ll use a dataset published by Allison and Cicchetti (1976). In this study, the authors studied the relationship between sleep and various ecological and morphological variables across a set of mammalian species: https://science.sciencemag.org/content/194/4266/732

Let’s start by loading the data into a table:

```{r}
allisontab <- tibble(read.csv("Data_allison.csv"))
```

This dataset contains several variables related to various body measurements and measures of sleep in different species. Note that some of these are continuous, while others are discrete and ordinal.

```{r}
summary(allisontab)
```

We'll start by comparing the body weight (`BodyWt`) and brain weight (`BrainWt`) measurements from all the species. As these are measurements that span multiple orders of magnitude, we'lll log-scale them before analyzing them. Later on, we'll also use the amount of time a species sleeps (`TotalSleep`) as well, so let's remove any rows that have missing data for any of these 3 variables.

```{r}
# Remove rows with missing data in columns of interest 
allisontab <- filter(allisontab,!is.na(BrainWt) & !is.na(BodyWt) & !is.na(TotalSleep))
# Log-scale body and brain weight
allisontab <- mutate(allisontab,logBody=log10(BodyWt), logBrain=log10(BrainWt))
```

Let's first look at the relationship between body weight and brain weight (now both in log-scale):

```{r}
plot(allisontab$logBrain,allisontab$logBody)
```

There seems to be a linear relationship here. But how confident are we in this? 

We'll use bootstrapping to assess the strength of this relationship: we'll build confidence intervals on the slope parameter of a linear regression of `logBrain` as a function of `logBody` using as few parametric assumptions as possible. The bootstrap comes to the rescue whenever we don't want to assume too much about the sampling distribution of our parameter. Below is a function to obtain a single bootstrapped sample from an input dataset. Take a close look at each step.

```{r}
bootstrap <- function(tab){
  # Preliminary check: if the table is a vector with a single variable, turn it into a matrix
  if(is.null(dim(tab))){tab <- matrix(tab,ncol=1)}
  # Count the number of elements in our data
  numelem <- nrow(tab)
  # Sample indexes with replacement
  bootsidx <- sample(1:numelem, replace=TRUE)
  # Obtain a boostrapped sample by selecting the bootstrapped indexes from the original sample
  final <- tab[bootsidx,]
  # Produce bootstrapped sample as output
  return(final)
}
```

Let's see what happens when we run this function on our data.

```{r}
boots <- bootstrap(allisontab)
plot(boots$logBrain,boots$logBody)
```

Repeat the above command lines multiple times. What happens?

Let's estimate a parameter: the slope coefficient in a linear regression of log brain weight on log body weight:

```{r}
lmmodel <- lm(logBrain ~ logBody, data=allisontab)
estimate <- lmmodel$coeff[2]
estimate
```

**Exercise**: try estimating the same parameter from a series of 1000 bootstrapped samples of our original data, and collecting each of the bootstrapped parameters into a vector called "bootsvec". Hint: you might want to use a for loop or a vectorized sapply() function.

```{r, echo=FALSE}
bootsvec <- c()
for(i in seq(1,1000)){
  boots <- bootstrap(allisontab)
  bootsest <- lm(logBrain ~ logBody, data=boots)$coeff[2]
  bootsvec <- c(bootsvec,bootsest)
}
# Optional: clean up the element names
names(bootsvec) <- c()
```

Let's plot the ecdf of all our estimates, using the function ecdf().

```{r}
plot(ecdf(bootsvec))
abline(v=estimate,col="red")
```

We are now ready to obtain confidence intervals (CIs) of our original parameter estimate, using our bootstrapped distribution. There are multiple ways to obtain CIs from a bootstrapped distribution. Some of these assume that the ECDF has particular properties, while others are more generally applicable: \
a) Standard error approach - assumes ECDF is approximately normal \
b) Percentile approach - assumes ECDF is symmetric and median-unbiased \
c) Pivotal approach - most general, makes almost no assumptions. \
These three approaches generally result in very similar CIs, but they differ (slightly) in methodology. In the interest of time, we'll demonstrate how to run the first two approaches in R. We'll leave the third approach as an exercise you can do at home (read Box 8-1 in the Edge book for an explanation of it, and a code example).

The standard approach is the most 'parametric' of the 3 approaches, and assumes that the ECDF is approximately a Normal distribution. Recall from the lecture on properties of estimators that we can obtain CIs from a Normally-distributed summary statistic $\theta$ by using the standard error of the statistic: $SE(\hat{\theta}_n)$:

$$(\hat{\theta} - SE(\hat{\theta}_n)q_{97.5\%}, \hat{\theta} + SE(\hat{\theta}_n)q_{97.5\%})$$
Where $q_{97.5\%}$ is a quantile from a standard N(0,1) Normal distribution, and $SE(\hat{\theta}_n)$ is the standard error (the standard deviation of our estimator). **The standard error approach to obtain boostrap confidence intervals proceeds by replacing the standard error $SE(\hat{\theta}_n)$ with the standard deviation of the bootstrap distribution as a stand-in for it.** Let's call this last value $SE_b(\hat{\theta}_n)$. Then:

$$(\hat{\theta} - SE_b(\hat{\theta}_n)q_{97.5\%}, \hat{\theta} + SE_b(\hat{\theta}_n)q_{97.5\%})$$
In R, we can thus obtain the CI around the mean as follows:

```{r}
# Let bootsvec be our vector of boostrapped parameters from the previous exercise
bootsmean <- mean(bootsvec)
SEb <- sd(bootsvec)
NormalQ <- qnorm(0.975,mean=0, sd=1)
CIs.bootsSE <- c( bootsmean - SEb*NormalQ, bootsmean + SEb*NormalQ )
names(CIs.bootsSE) <- c("2.5%","97.5%")
CIs.bootsSE
```

Note that here, we are still using Normal quantiles, which is why this method has a bit of a parametric flavor. 

The percentile approach is less 'parametric' and simply involves obtaining percentiles of the bootstrap distribution. If we are interested in a 95% CI, for example, we just look for the values of the bootstrap distribution that correspond to the 2.5% percentile and the 97.5% percentile, so that the central bootstrap distribution mass between them adds up to 95%.

**Exercise**: To obtain the bootstrap-based quantile confidence intervals, compute the 95% quantiles (lower 2.5% and upper 97.5%) of the bootstrap distribution we've obtained before.

```{r, echo=FALSE, eval=FALSE}
# Let bootsvec be our vector of boostrapped parameters from the previous exercise
CIs.bootsQ <- quantile(bootsvec, c(0.025,0.975))
CIs.bootsQ
```

Compare these confidence intervals to the (highly parametric) CIs that are obtained directly from the linear model, using the function `confint`, which assumes that the fitted parameter is normally distributed, and does not use any information about bootstraps at all (just the standard error of the fitted model).


```{r}
CIs.param <- confint(lmmodel)[2,]
CIs.param
```

> While it is easy to obtain parametric CIs in this case (as this is a simple linear model), it will be much harder to obtain parametric CIs for more complex models where normality assumptions are harder to justify. Bootstrap-based CIs, in contrast, are obtainable regardless of how complex your model is: you just need to bootstrap your data many times and fit you your model on each bootstraped sample!

## Permutation test

Let's evaluate the relationship that there is no relationship between logBrain and logBody. Recall that one way to do it would be by using a linear model, and testing whether the value of the fitted slope is significantly different from zero, using a t-test:

```{r}
summary(lm(logBrain ~ logBody, data=allisontab))
```

This test, however, makes assumptions on our data that sometimes may not be warranted, like large sample sizes and homogeneity of variance. We can perform a more general test that makes less a priori assumptions on our data - a permutation test - as long as we are careful in permuting the appropriate variables for the relationship we are trying to test. In this case, we only have two variables, and we are trying to test whether there is a significant relationship between them. If we randomly shuffle one variable with respect to the other, we should obtain a randomized sample of our data. We can use the following function, which takes in a tibble and a variable of interest, and returns a new tibble in which that particular variable's values are randomly shuffled.

```{r}
permute <- function(tab,vartoshuffle){
  # Extract column we wish to shuffle as a vector
  toshuffle <- unlist(tab[,vartoshuffle],use.names=FALSE)
  # The function sample() serves to randomize the order of elements in a vector
  shuffled <- sample(toshuffle)
  # Replace vector in new table (use !! to refer to a dynamic variable name)
  newtab <- mutate(tab, !!vartoshuffle := shuffled )
  return(newtab)
}
```

Now we can obtain a permuted version of our original data, and compute the slope estimate on this dataset instead:

```{r}
permuted <- permute(allisontab, "logBrain")
plot(permuted$logBody,permuted$logBrain)
permest <- lm(logBrain ~ logBody, data=permuted)$coeff[2]
permest
```

**Exercise**: try estimating the same parameter from a series of 100 permuted versions of our original data, and collecting each of the permuted parameters into a vector called "permvec".


```{r, echo=FALSE}
permvec <- c()
for(i in seq(1,100)){
  permdat <- permute(allisontab,"logBrain")
  colnames(permdat) <- c("Species","logBody","logBrain")
  permest <- lm(logBrain ~ logBody, data=permdat)$coeff[2]
  permvec <- c(permvec,permest)
}
# Optional: clean up the element names
names(permvec) <- c()
```

We now have a distribution of the parameter estimate under the assumption that there is no relationship between these two variables:

**Exercise**: obtain an empirical one-tailed P-value from this distribution by counting how many of the permuted samples are as large as our original estimate, and dividing by the total number of permuted samples we have.


```{r, echo=FALSE}
data_est <- lm(TotalSleep ~ log(BodyWt), data=allisontab)$coeff[2]
approx_pval <- (sum(permvec < data_est) + 1)/(length(permvec) + 1)
```


## Validation

We'll perform a validation exercise to evaluate the error of various models on the data. In this case, we'll create a predictor for TotalSleep as a function of logBody, using a linear model, and then test how well it does. We'll first divide our data into a "training" partition - which we'll use to fit our model - and a separate "test" partition - which we'll use to test how well our model is doing, and avoid over-fitting. Each partition will be one half of our original data.

```{r}
# Obtain the number of data points we have
numdat <- dim(allisontab)[1]
# For the training set, randomly sample 50% of the data indexes
trainset <- sample(numdat, round(numdat*0.5))
# For the test set, obtain all indexes that are not in training set
testset <- seq(1,numdat)[-trainset]
```

Let's begin by calculating the mean squared error (MSE) between our observatiosn and our predictions in our test partition, after fitting a linear model to our training partition:

```{r}
# Fit the linear model to the training subset of the data
fit1 <- lm(TotalSleep ~ logBody, data=allisontab,subset=trainset)
# Predict all observations using the fitted linear model
predall <- predict(fit1,allisontab)
# Compute mean squared differences between observations and predictions
sqdiff <- (allisontab$logBrain - predall)^2 
# Extract the differences for the test partition
sqdiff.test <- sqdiff[testset]
# Compute the mean squared error
mse1 <- mean(sqdiff.test)
```

Now, we'll try to fit our data to two more complex models: a quadratic model and a cubic model, using the function `poly`:

```{r}
fit2 <- lm(TotalSleep ~ poly(logBody,2), data=allisontab,subset=trainset)
mse2 <- mean(((allisontab$logBrain - predict(fit2,allisontab))^2)[testset])

fit3 <- lm(TotalSleep ~ poly(logBody,3), data=allisontab,subset=trainset)
mse3 <- mean(((allisontab$logBrain - predict(fit3,allisontab))^2)[testset])
```

We can see that the MSE appears to increase for the more complex models. This suggests a simple linear fit performs better at predicting values that were not included in the training set.

**Exercise**: compute the MSE on the training partition. Compare the resulting values to the MSE on the test partition. What do you observe? Is the difference in errors between the three models as large as when computing the MSE on the test partition? Why do you think this is?

```{r, echo=FALSE, eval=FALSE}
mse2.training <- mean(((allisontab$logBrain - predict(fit2,allisontab))^2)[trainset])
mse3.training <- mean(((allisontab$logBrain - predict(fit2,allisontab))^2)[trainset])
```

## Cross-validation

We'll now perform a cross-validation exercise. If you haven't installed it, you'll need to install the library `boot` before loading it. This allows you to perform various resampling techniques, including both bootstrapping and cross-validation.

```{r}
if(require("boot") == FALSE){install.packages("boot")}
library("boot")
```

The function `cv.glm()` from the library `boot` can be used to compute a cross-validation error. This function is designed to work with the `glm()` function for fitting generalized linear models in R, but we can compute a simple linear regression using `glm()` as well, and then feed the result into `cv.glm()`:

```{r}
fit1=glm( TotalSleep ~ logBody, data=allisontab )
# The LOOCV error is computed using the function cv.glm()
cv.err=cv.glm(allisontab,fit1)
```

The value of the cross-validation error is stored in the second element of the attribute `delta` of the output of `cv.glm`. By default, this is a "leave-one-out" cross-validation (LOOCV) error, meaning it computes error by leaving 1 data point out of the fitting and evaluating the error at that data point. The process is iterated over all data points, and the errors are then averaged together. We can obtain the value of the LOOCV error by writing:

```{r}
cv.err$delta[1]
```

Now, let's compute the LOOCV error for increasingly complex polynomial models (linear, quadratic, cubic, etc.):

```{r}
CVerr=rep(0,5)
for(m in 1:5){
  fit=glm(TotalSleep ~ poly(logBody,m), data=allisontab)
  CVerr[m]=cv.glm(allisontab,fit)$delta[1]
}
```

**Exercise**: Plot the results of this cross-validation exercise. Which model has the lowest LOOCV error?

```{r, echo=FALSE, eval=FALSE}
plot(seq(1,5),CVerr,xlab="Model complexity",ylab="LOOCV error")
```

The LOOCV error is approximately unbiased for the true prediction error, but often has high variance, and can be computationally expensive when our datasets are large. 5-fold and 10-fold cross-validation are recommended as a good compromise between reduced variance and a bit of bias in the estimation of the prediction error. Let's try increasing the size of the K-fold partition.

**Exercise**: Take a look at the help function for `cv.glm`. Which argument would you modify to be able to compute the 10-fold cross-validation error, instead of the LOOCV error. Can you do this for the five models we tested above? Which model has the lowest cross-validation error now?

```{r, echo=FALSE, eval=FALSE}
CVK10err=rep(0,5)
for(m in 1:5){
  fit=glm(TotalSleep ~ poly(logBody,m), data=allisontab)
  CVK10err[m]=cv.glm(allisontab,fit,K=10)$delta[1]
}
plot(seq(1,5),CVK10err,xlab="Model complexity",ylab="10-fold CV error")
```
